Practical Effects of Integrating Temperature with Strang Split Reactions

### 2021-03-22 strang/paper.tex
word count

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
For astrophysical reacting flows, operator splitting is commonly used
to couple hydrodynamics and reactions.  Each process operates
independent of one another, but by staggering the updates in a
symmetric fashion (via Strang splitting) second order accuracy in time
can be achieved.  However, approximations are often made to the
reacting system, including the choice of whether or not to integrate
temperature with the species.  Here we demonstrate through a simple
convergence test that integrating an energy equation together with reactions
achieves the best convergence when modeling reactive flows with Strang
splitting.
\end{abstract}

@@

Simulations of stellar flows require solving the equations of
hydrodynamics coupled with a nuclear reaction network.  The equations
of hydrodynamics with reacting sources are:
\begin{align}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &= 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho X_k \Ub) &= \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &= 0 \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot \left [ (\rho E + p) \Ub \right ] &= \rho \Sdot
\end{align}
where $\rho$ is the density, $\Ub$ is the velocity vector, $X_k$ are the species mass fractions
with creation rates $\omegadot_k$ from reactions, $p$ is the pressure, $E$ is the
specific total energy, and $\Sdot$ is the nuclear energy generation rate.

When we are reacting, we can look at internal energy
\begin{equation}
\rho \frac{De}{Dt} + p \nabla \cdot \Ub = \rho \Sdot,
\end{equation}
where $e$ is the specific internal energy or alternately, we can evolve the temperature, $T$,
via
\begin{equation}
\rho c_v \frac{DT}{Dt} = \rho \left (\frac{p}{\rho^2} - \left .\frac{\partial e}{\partial \rho} \right |_T \right ) \frac{D\rho}{Dt} + \rho \Sdot,
\end{equation}
where $c_v = \partial e / \partial T |_\rho$ (this form neglects composition
changes, see \citealt{ABNZ:III}, which assumes chemical
equilibrium).

@@

In \citet{castro_sdc}, we introduced a test problem where we could
measure the convergence of a reacting hydrodynamics problem via
Richardson extrapolation---this was an acoustic pulse with helium
burning via $3$-$\alpha$ and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$.  Initially, the domain is
pure $\isotm{He}{4}$, but both $\isotm{C}{12}$ and $\isotm{O}{16}$ are
created as time evolves.  The published tests focused on SDC coupling
and showed that we can get overall 4th order in space and time
convergence.  Here we run the same test with Strang coupling, looking
at the various approaches at incorporating a temperature / energy equation in
the reactive portion of the update.

@@

Figure~\ref{thefigure} shows the norm of the error vs.\ the coarse run
resolution.  The slope of these lines is a measure of the convergence
rate~\citep{ORVV}.  Guidelines show ideal first and second order
convergence.  We see that all methods converge at least second order
for density, but for the thermodynamic quantities, $(\rho e)$ and $T$,
the method where only $X_k$ is evolved during reactions has larger
errors and much worse convergence than the other methods.  Looking at
the trace nuclei generated in the burning, $\isotm{C}{12}$ and
$\isotm{O}{16}$, we see a large difference between the two methods
that evolve some sort of energy and the one method where only $X_k$ is
evolved---the latter converging essentially first order at high
resolution.

>>>>

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
For astrophysical reacting flows, operator splitting is commonly used
to couple hydrodynamics and reactions.  Each process operates
independent of one another, but by staggering the updates in a
symmetric fashion (via Strang splitting) second order accuracy in time
can be achieved.  However, approximations are often made to the
reacting system, including the choice of whether or not to integrate
temperature with the species.  Here we demonstrate through a simple
convergence test that integrating an energy equation together with reactions
achieves the best convergence when modeling reactive flows with Strang
splitting.  Additionally, second order convergence cannot be achieved without integrating an energy or temperature equation.
\end{abstract}

@@

Simulations of stellar flows require solving the equations of
hydrodynamics coupled with a nuclear reaction network.  The equations
of hydrodynamics with reacting sources are:
\begin{align}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &= 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho X_k \Ub) &= \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &= 0 \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot \left [ (\rho E + p) \Ub \right ] &= \rho \Sdot
\end{align}
where $\rho$ is the density, $\Ub$ is the velocity vector, $X_k$ are the species mass fractions
with creation rates $\omegadot_k$, $p$ is the pressure, $E$ is the
specific total energy, and $\Sdot$ is the nuclear energy generation rate.

When we are reacting, we can look at internal energy
\begin{equation}
\rho \frac{De}{Dt} + p \nabla \cdot \Ub = \rho \Sdot,
\end{equation}
where $e$ is the specific internal energy or alternately, we can evolve the temperature, $T$,
\begin{equation}
\rho c_v \frac{DT}{Dt} = \rho \left (\frac{p}{\rho^2} - \left .\frac{\partial e}{\partial \rho} \right |_T \right ) \frac{D\rho}{Dt} + \rho \Sdot,
\end{equation}
where $c_v = \partial e / \partial T |_\rho$ (this form neglects composition
changes, see \citealt{ABNZ:III}).

@@

In \citet{castro_sdc}, we introduced a test problem where we could
measure the convergence of a reacting hydrodynamics problem via
Richardson extrapolation---this was an acoustic pulse with helium
burning via $3$-$\alpha$ and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$.  Initially, the domain is
pure $\isotm{He}{4}$, but both $\isotm{C}{12}$ and $\isotm{O}{16}$ are
created as time evolves.  The published tests 
showed that we can get overall 4th order in space and time
convergence with SDC coupling.  Here we run the same test with Strang coupling, looking
at the various approaches at incorporating a temperature / energy equation in
the reactive portion of the update.

@@

Figure~\ref{thefigure} shows the norm of the error vs.\ the coarse run
resolution.  The slope of these lines is a measure of the convergence
rate~\citep{ORVV}.  We see that all methods converge at least second order
for density, but for the thermodynamic quantities, $(\rho e)$ and $T$,
the method where only $X_k$ is evolved during reactions has larger
errors and much worse convergence than the other methods.  Looking at
the trace nuclei generated in the burning, $\isotm{C}{12}$ and
$\isotm{O}{16}$, we see a large difference between the two methods
that evolve some sort of energy and the one method where only $X_k$ is
evolved---the latter converging essentially first order at high
resolution.
